### 2014-07-09 Paper_v1.tex
going home ...

The rule based regulatory strategy captured classic patterns of allosteric activation and inhibition of enzyme activity (Fig. \ref{fig-kinetics-simulations}).  
The first case we explored was substrate activation of enzyme activity (positive cooperativity). 
Classically, we expected a sigmoidal relationship substrate abundance and reaction rate for positive cooperativity.

@@

\subsection*{Formulation and solution of the model equations.}
We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme ($\epsilon_{i}$) abundance in hypothetical cell free
metabolic networks:
\begin{eqnarray}
	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of
enzymes in the model. The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ 
denotes the rate of reaction $j$. Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters 
$\mathbf{k}$ ($\mathcal{K}\times{1}$).
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
$\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
metabolite $i$ is not connected with reaction $j$. 
Each reaction rate was written as the product of a reaction term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
\begin{equation}
	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
\end{equation}We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
\begin{equation}\label{eqn:rate-bar}
	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$). 
The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
For each enzyme, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
If an enzyme was activated by $m$ metabolites, we modeled this activation as:
\begin{equation}
	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
\end{equation}Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
\begin{equation}
	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
\begin{equation}
	v_{j} = \min\left(u_{j},d_{j}\right)
\end{equation}where:
\begin{eqnarray}
	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right) \\
	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
\end{eqnarray}where $j^{+}$ and $j^{-}$ denote the sets of activating, and inhibitory factors for enzyme $j$. 
If an enzyme had no allosteric factors, we set $v_{j} = 1$.
In this study, each individual factor had the form:
\begin{equation}
	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
\end{equation}where $x_{j}$ denotes the abundance of metabolite $j$, and $\kappa_{ij}$ and $\eta$ are control parameters. 
The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave [REFHERE].

@@

\begin{figure}
\centering
\includegraphics[width=0.55\textwidth]{./figs/Figure-3-OnOffSimulations.pdf}
\caption{On/off control simulations for network A and network B for an ensemble of kinetic parameter sets. 
For each case, N = 100 simulations were conducted using kinetic and initial conditions randomly generated from a hypothetical true parameter set. 
The gray area represents $\pm$ one standard deviation surrounding the mean. Control parameters were fixed during the ensemble calculations.}\label{fig-onoff-simulations}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-4-ParameterFit.pdf}
\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic. }\label{fig-parameter-fit}
\end{figure}

>>>>

The rule based regulatory strategy captured the kinetic behavior of classic allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}).  
We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
Consistent with classical data, the rule based strategy predicted a sigmoidal relationship between substrate abundance 
and reaction rate as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}A). 
For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. This was consistent with the idea that
substrate binding \textit{decreases} at regulatory sites negatively impacts the ability of the enzyme to bind substrate at the active site. 
On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate $S$ to product $P$ by enzyme $E$ approached a step function.
In the presence of an inhibitor, the rule based strategy predicted non-competitive like behavior as a function of the 
cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
When the control gain parameter, $\kappa_{ij}$ in Eqn. \eqref{eqn:control-factor}, was greater than unity, the inhibitory force was
directly proportional to the cooperativity parameter, $\eta$ in Eqn. \eqref{eqn:control-factor}. 
Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B, orange).
However, when the gain parameter was less than unity, enzyme inhibition increased with \textit{decreasing} cooperativity, 
i.e., smaller $\eta$ yielded increased inhibition (Fig. \ref{fig-kinetics-simulations}B).
Interestingly, our rule based approach was not able to directly simulate competitive inhibition of enzyme activity. 
Taken together, the rule based strategy captured classical regulatory patterns for both enzyme activation and inhibition.

@@

\subsection*{Formulation and solution of the model equations.}
We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme ($\epsilon_{i}$) abundance in hypothetical cell free
metabolic networks:
\begin{eqnarray}
	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of
enzymes in the model. The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ 
denotes the rate of reaction $j$. Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters 
$\mathbf{k}$ ($\mathcal{K}\times{1}$).
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. If 
$\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates
metabolite $i$ is not connected with reaction $j$. 
Each reaction rate was written as the product of a reaction term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
\begin{equation}
	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
\end{equation}We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
\begin{equation}\label{eqn:rate-bar}
	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$). 
The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
For each enzyme, we used a rule based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
If an enzyme was activated by $m$ metabolites, we modeled this activation as:
\begin{equation}
	v_{j} = \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
\end{equation}Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
\begin{equation}
	v_{j} = 1 - \max\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right)
\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
\begin{equation}
	v_{j} = \min\left(u_{j},d_{j}\right)
\end{equation}where:
\begin{eqnarray}
	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(x\right),\hdots,f_{mj}\left(x\right)\right) \\
	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(x\right),\hdots,f_{nj}\left(x\right)\right)
\end{eqnarray}where $j^{+}$ and $j^{-}$ denote the sets of activating, and inhibitory factors for enzyme $j$. 
If an enzyme had no allosteric factors, we set $v_{j} = 1$.
In this study, each individual factor had the form:
\begin{equation}\label{eqn:control-factor}
	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}x_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}x_{j}^{\eta}}
\end{equation}where $x_{j}$ denotes the abundance of metabolite $j$, and $\kappa_{ij}$ and $\eta$ are control parameters. 
The $\kappa_{ij}$ parameter was species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).
The model equations were encoded using the Octave programming language, and solved using the LSODE routine in Octave [REFHERE].

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-3-EnzymeKinetics.pdf}
\caption{Kinetics of simple transformations in the presence of activation and inhibition. 
A:The conversion of substrate $S$ to product $P$ by enzyme $E$ was activated by $S$. 
B:The conversion of substrate $S$ to product $P$ by enzyme $E$ was inhibited by inhibitor $I$. 
}\label{fig-kinetics-simulations}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.55\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
\caption{On/off control simulations for network A and network B for an ensemble of kinetic parameter sets versus time. 
For each case, N = 100 simulations were conducted using kinetic and initial conditions randomly generated from a hypothetical true parameter set. 
The gray area represents $\pm$ one standard deviation surrounding the mean. Control parameters were fixed during the ensemble calculations.}\label{fig-onoff-simulations}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic. }\label{fig-parameter-fit}
\end{figure}
